---
title: "analyses feather structure"
author: ""
date: "2026-06-06"
output: html_document
---

```{r}
library(readr)
library(readxl)
library(dplyr)
library(ggplot2)
library(lmerTest)
library(emmeans)
library(ggeffects)
library(multcomp)
library(cowplot)
```

#0. Import and clean data
```{r}
FeatherAd_db <- read_csv("feather.csv")
```

#1. Are there differences in the feather microstructure between the urban and rural habitat
##1.1 Model
```{r}
FeatherAd_db$area_habitat<-paste(FeatherAd_db$area, FeatherAd_db$habitat, sep="_")

#change names
FeatherAd_db <- FeatherAd_db %>%
  mutate(habitat = factor(habitat, levels = c("Urban", "Forest"), labels = c("Urban", "Rural")))

model_habitat <- lmer(mean_density ~ habitat + sex + age + habitat*age + (1|area_habitat)+ (1|area) +(1|id), data=FeatherAd_db) 
summary(model_habitat)
anova(model_habitat)

plot(predict_response(model_habitat, terms="habitat"))

#some tests
hist(residuals(model_habitat))
qqnorm(resid(model_habitat))
qqline(resid(model_habitat), col = "red")

model_habitat_2 <- lmer(mean_density ~ habitat + sex + age + (1|area_habitat)+(1|area) +(1|id), data=FeatherAd_db) 
summary(model_habitat_2)
anova(model_habitat_2)

#some tests
hist(residuals(model_habitat_2))
qqnorm(resid(model_habitat_2))
qqline(resid(model_habitat_2), col = "red")
```

##1.2 Plot
```{r}
pred_general <- as.data.frame(predict_response(model_habitat, terms = "habitat"))

p1 <- ggplot() +
  # --- Jittered raw data points ---
  geom_jitter(
    data = FeatherAd_db,
    aes(x = habitat, y = mean_density, colour = habitat),
    width  = 0.15,
    height = 0,
    size   = 1.8,
    alpha  = 0.35,
    shape  = 19
  ) +
    # --- Confidence interval error bars ---
  geom_errorbar(
    data = pred_general,
    aes(x = x, ymin = predicted-std.error, ymax = predicted+std.error),
    colour="black",
    width     = 0.08,
    linewidth = 1.0
  ) +
  # --- Predicted mean points ---
  geom_point(
    data = pred_general,
    aes(x = x, y = predicted, fill = x),
    size  = 5,
    shape = 21,
    stroke = 1.5
  ) +

  # --- Colours for Forest vs Urban ---
  scale_fill_manual(
    values = c(
      "Rural" = "springgreen4",
      "Urban"  = "grey50"
    )
  ) +
    scale_color_manual(
    values = c(
      "Rural" = "springgreen4",
      "Urban"  = "grey50"
    )
  ) +

  labs(
    x      = "Habitat",
    y      = "Mean barbule density (barbules/mm)"
  ) +

  theme_bw(base_size = 20) +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(size = 14, face = "bold"),
    axis.title.x    = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
  )+
  scale_y_continuous(
  limits = c(14.9, 27),           # min and max of y axis
  breaks = seq(16, 26, by = 2)  # breaks every 2 units: 16, 18, 20, 22, 24, 26
)
  
p1

#ggsave("per_pop.png", plot = p1, width = 8, height = 7, dpi = 600)
```

#2. Are the urban effects consisten across space
##2.1 Model
```{r}
model_space <- lmer(mean_density ~ habitat*area + sex + age + (1|id), data=FeatherAd_db) 
summary(model_space)
anova(model_space) 

#the interaction is not significant, meaning the (lack of) effects of habitat are consistent in space
FeatherAd_db2<-FeatherAd_db
FeatherAd_db2$area <- factor(FeatherAd_db2$area,
                             levels = c("Lisbon", "Madrid", "Milan", "Malmo", "Goteborg"),
                             labels = c("Lisbon", "Madrid", "Milan", "Malmo", "Gothenburg"))


model_space_2 <- lmer(mean_density ~ habitat+area + sex + age + (1|id), data=FeatherAd_db2) 
summary(model_space_2)
anova(model_space_2) 
```

###2.1.1 Post hoc test
```{r}
#Marginal means per area
emm_area <- emmeans(model_space_2, ~ area)
emm_area

#Pairwise comparisons
pairs_area <- pairs(emm_area, adjust = "tukey")
summary(pairs_area)
confint(pairs_area)

#Put letters for the differences beween areas
library(multcomp)
cld(emm_area, adjust = "tukey", Letters = letters)

#Letters
cld_df <- as.data.frame(cld(emm_area, adjust = "tukey", Letters = letters))
cld_df$.group <- trimws(cld_df$.group)  # remove whitespace from letters
```

### 2.2.2 Plot
```{r}
# Order the pops.
lat_order <- c("Lisbon", "Madrid", "Milan", "Malmo", "Gothenburg")

FeatherAd_db2$area  <- factor(FeatherAd_db2$area,  levels = lat_order)
cld_df$area        <- factor(cld_df$area,        levels = lat_order)

#Model predictions
pred_space     <- predict_response(model_space_2, terms = c("area", "habitat"))
pred_space_df  <- as.data.frame(pred_space)
pred_space_df$x <- factor(pred_space_df$x, levels = lat_order)

#Label position
label_pos <- pred_space_df %>%
  group_by(x) %>%
  summarise(label_y = max(predicted + std.error) + 0.3)

cld_df <- cld_df %>%
  left_join(label_pos, by = c("area" = "x"))

#Color
hab_colours <- c("Rural" = "springgreen4", "Urban" = "grey50")

p2 <- ggplot() +
  # --- Raw jittered points ---
  geom_jitter(
    data   = FeatherAd_db2,
    aes(x = area, y = mean_density, colour = habitat),
    width  = 0.2,
    height = 0,
    size   = 1.5,
    alpha  = 0.3,
    shape  = 16
  ) +
  # --- SE error bars ---
  geom_errorbar(
    data = pred_space_df,
    aes(x = x, ymin = predicted - std.error, ymax = predicted + std.error,
        colour = group),
    width     = 0.1,
    linewidth = 0.9,
    position  = position_dodge(width = 0.5)
  ) +
  # --- Predicted means ---
  geom_point(
    data = pred_space_df,
    aes(x = x, y = predicted, fill = group),
    size     = 4,
    shape    = 21,
    stroke   = 1.2,
    position = position_dodge(width = 0.5)
  ) +
  # --- CLD letters ---
  geom_text(
    data    = cld_df,
    aes(x   = area, label = .group),
    y       = 14.7,
    size    = 4,
    colour  = "black"
  ) +
  scale_y_continuous(
    limits = c(14.9, 27),
    breaks = seq(16, 26, by = 2)
  ) +
  scale_fill_manual(values   = hab_colours) +
  scale_colour_manual(values = hab_colours) +
  labs(
    y      = "",
    fill   = "Habitat",
    colour = "Habitat"
  ) +
    labs(
    x      = "Area",
    y      = "Mean barbule density (barbules/mm)"
  ) +
  theme_bw(base_size = 20) +
  theme(
    legend.position    = "none",
    axis.title.x       = element_blank(),
    axis.text.x     = element_text(size = 14, face = "bold"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
  )

p2

#ggsave("Plots brevia/per_pop.png", plot = p2, width = 8, height = 7, dpi = 600)
```

#4. Exploring the effects of latitude, temperature and elevation
##4.1 add elevation, temperature and latitude
```{r}
elevation <- read.csv("elevation.csv")

elevation2 <- elevation %>% 
  mutate(habitat = area,
         area = city)

elevation2 <- elevation2 %>% 
  dplyr::select(habitat, area, elevation, latitude)

FeatherAd_db2$latitude <- NULL
FeatherAd_db2 <- left_join(FeatherAd_db2, elevation2, by = c("habitat", "area"))
FeatherAd_db2$elevation <- as.numeric(FeatherAd_db2$elevation)

# sanity check — should return 0 rows if the join worked fully
anti_join(FeatherAd_db2, elevation2, by = c("habitat", "area")) %>% nrow()

#tmean
tmean_long_filt <- read_csv("tmean_long_filt.csv")
tmean_long_filt$...1 <- NULL

# same reasoning — removed the two ifelse() recodes that reintroduced
# "Goteborg" and "Forest", which no longer match FeatherAd_db2
tmean_long_filt <- tmean_long_filt %>% 
  rename(habitat = area,
         area = city)

FeatherAd_db2 <- left_join(FeatherAd_db2, tmean_long_filt, by = c("habitat", "area"))

# sanity check
anti_join(FeatherAd_db2, tmean_long_filt, by = c("habitat", "area")) %>% nrow()
```

##4.2 Model latitude
```{r}
model_latitude <- lmer(mean_density ~  sex + (latitude) + (1|area_habitat)+ (1|area) +(1|id), data=FeatherAd_db2) 
summary(model_latitude)
anova(model_latitude) #No significant effects
```

##4.2 Model elevation
```{r}
model_elevation <- lmer(mean_density ~  sex + age + elevation + (1|area_habitat) + (1|area) +(1|id), data=FeatherAd_db2) 
summary(model_elevation)
anova(model_elevation) 

# summarise mean ± SE per area_habitat (elevation is constant within each)
summary_db <- FeatherAd_db2 %>%
  group_by(area_habitat, elevation) %>%
  summarise(mean_md = mean(mean_density, na.rm = TRUE),
            se_md = sd(mean_density, na.rm = TRUE) / sqrt(n()),
            .groups = "drop")

pred_elev <- ggpredict(model_elevation, terms = "elevation [n=100]")

plot_elevation<-ggplot() +
  geom_ribbon(data = pred_elev, aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "grey40") +
  geom_line(data = pred_elev, aes(x = x, y = predicted),
            color = "black", linewidth = 1, linetype="solid") +
  geom_errorbar(data = summary_db, aes(x = elevation, ymin = mean_md - se_md, ymax = mean_md + se_md),
                width = 0, color = "black") +
  geom_point(data = summary_db, aes(x = elevation, y = mean_md),
             color = "black", size = 3) +
  labs(x = "Elevation (m)", y = "Mean barbule density (barbule/mm)") +
  scale_y_continuous(
  limits = c(19.3, 22.1),           # min and max of y axis
  breaks = seq(19.5, 22, by = 0.5)  # breaks every 2 units: 16, 18, 20, 22, 24, 26
)+
  theme_bw(base_size = 20) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank(),
  )

plot_elevation

#ggsave("plot_elevation.png", plot = plot_elevation, width = 8, height = 7, dpi = 600)
```

```{r}
base_colors <- c(
  "Madrid"   = "#E63946",
  "Lisbon"   = "#2A9D8F",
  "Milan"    = "#457B9D",
  "Malmo"    = "#F4A261",
  "Gothenburg" = "#8E44AD"
)

summary_db <- FeatherAd_db2 %>%
  group_by(area, habitat, elevation) %>%
  summarise(mean_md = mean(mean_density, na.rm = TRUE),
            se_md = sd(mean_density, na.rm = TRUE) / sqrt(n()),
            .groups = "drop") %>%
  mutate(fill_color = ifelse(habitat == "Urban", base_colors[area], "white"))

plot_elevation <- ggplot() +
  geom_ribbon(data = pred_elev, aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "grey40") +
  geom_line(data = pred_elev, aes(x = x, y = predicted),
            color = "black", linewidth = 1, linetype = "solid") +
  geom_errorbar(data = summary_db, aes(x = elevation, ymin = mean_md - se_md, ymax = mean_md + se_md,
                                        color = area),
                width = 0) +
  geom_point(data = summary_db, aes(x = elevation, y = mean_md, color = area, fill = fill_color),
             shape = 21, size = 4, stroke = 1.2) +
  scale_color_manual(values = base_colors, name = "Area") +
  scale_fill_identity() +
  labs(x = "Elevation (m)", y = "Mean barbule density (barbules/mm)") +
  scale_y_continuous(
    limits = c(19.3, 22.1),
    breaks = seq(19.5, 22, by = 0.5)
  ) +
  theme_bw(base_size = 20) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank()
  )

plot_elevation
#ggsave("plot_elevation.png", plot = plot_elevation, width = 8, height = 7, dpi = 600)
```

```{r}
library(cowplot)

base_colors <- c(
  "Madrid"   = "#E63946",
  "Lisbon"   = "#2A9D8F",
  "Milan"    = "#457B9D",
  "Malmö"    = "#F4A261",
  "Gothenburg" = "#8E44AD"
)

FeatherAd_db2 <- FeatherAd_db2 %>%
  mutate(area = recode(area, "Malmo" = "Malmö"))

summary_db <- FeatherAd_db2 %>%
  group_by(area, habitat, elevation) %>%
  summarise(mean_md = mean(mean_density, na.rm = TRUE),
            se_md = sd(mean_density, na.rm = TRUE) / sqrt(n()),
            .groups = "drop") %>%
  mutate(fill_color = ifelse(habitat == "Urban", base_colors[area], "white"))

# Your original working plot, just re-enable the color legend
plot_elevation <- ggplot() +
  geom_ribbon(data = pred_elev, aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.1, fill = "grey40") +
  geom_line(data = pred_elev, aes(x = x, y = predicted),
            color = "black", linewidth = 1, linetype = "solid") +
  geom_errorbar(data = summary_db, aes(x = elevation, ymin = mean_md - se_md, ymax = mean_md + se_md,
                                        color = area),
                width = 0) +
  geom_point(data = summary_db, aes(x = elevation, y = mean_md, color = area, fill = fill_color),
             shape = 21, size = 4, stroke = 1.2) +
  scale_color_manual(values = base_colors, name = NULL) +
  scale_fill_identity() +
  labs(x = "Elevation (m)", y = "Mean barbule density (barbules/mm)") +
  scale_y_continuous(limits = c(19.3, 22.1), breaks = seq(19.5, 22, by = 0.5)) +
  theme_bw(base_size = 20) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 12),  # <-- legend size
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
guides(
  color = guide_legend(
    nrow = 1,
    override.aes = list(
      shape = 21, 
      size = 2.5, 
      fill = base_colors,
      color = base_colors  # border same as fill
    )
  )
)

# Separate mini plot just for the habitat key
habitat_legend <- ggplot(data.frame(habitat = c("Urban", "Rural"),
                                    x = c(3.1, 3.9), y = c(1, 1))) +
  geom_point(aes(x = x, y = y, fill = habitat), 
             shape = 21, size = 2.5, stroke = 1.2, color = "black") +
  geom_text(aes(x = x, y = y, label = habitat), 
            hjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Urban" = "black", "Rural" = "white")) +
  theme_void() +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  xlim(0, 5) +
  ylim(0.5, 1.5)

p2<-plot_grid(plot_elevation, habitat_legend, ncol = 1, rel_heights = c(1, 0.12))

p2
#ggsave("p2.pdf", plot = p2, width = 7, height = 7, dpi = 600)
```




##4.2 Model temperature
```{r}
tmean_long <- read_csv("Eobs for salmon/tmean_long.csv")
str(tmean_long)

library(lubridate)

monthly_means <- tmean_long %>%
  filter(month(Date) %in% 5:8) %>% #get relevant months
  mutate(month = month(Date, label = TRUE, abbr = TRUE)) %>%
  group_by(city, area) %>%
  summarise(mean_Tmean = mean(Tmean, na.rm = TRUE), .groups = "drop") %>% 
  mutate(city = if_else(city == "Malmo", "Malmö", city))

FeatherAd_db2<-left_join(FeatherAd_db2, monthly_means, by=c("habitat"="area", "area"="city"))

model_temp <- lmer(mean_density ~ sex + age + mean_Tmean + (1|area_habitat)+ (1|area) +(1|id), data=FeatherAd_db2) 
summary(model_temp)
anova(model_temp) #marginal effects
performance::check_model(model_temp)


str(FeatherAd_db2)
```

##4.3 Solar radiation
```{r}
UV_values_annual_May_AU <- read_csv("UV_values_annual_&_May_AU.csv")

UV_values_annual_May_AU<-UV_values_annual_May_AU %>% 
  mutate(city = if_else(city == "Malmo", "Malmö", city))


FeatherAd_db2_UV<-left_join(FeatherAd_db2, UV_values_annual_May_AU, by=c("area"="city", "habitat"="area"))

##May-Aug
model_UV_mayau <- lmer(mean_density ~ sex + age + UV_B_may_au + an_tmean + (1|area_habitat)+ (1|area) +(1|id), data=FeatherAd_db2_UV) 
summary(model_UV_mayau) 
```

```{r}
library(lme4)
library(lmerTest)
library(ggeffects)  
library(ggplot2)
library(dplyr)

pred <- ggpredict(model_UV_mayau, terms = "UV_B_may_au [n=100]")

raw_summary <- FeatherAd_db2_UV %>%
  group_by(area_habitat) %>%
  summarise(
    UV_mean = mean(UV_B_may_au, na.rm = TRUE),
    mean_dens = mean(mean_density, na.rm = TRUE),
    se_dens = sd(mean_density, na.rm = TRUE) / sqrt(n()),
    n = n()
  )

ggplot() +
  geom_ribbon(data = pred, aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.2, fill = "steelblue") +
  geom_line(data = pred, aes(x = x, y = predicted),
            color = "steelblue", linewidth = 1) +
  geom_point(data = raw_summary, aes(x = UV_mean, y = mean_dens),
             color = "black", size = 2) +
  geom_errorbar(data = raw_summary,
                aes(x = UV_mean, ymin = mean_dens - se_dens, ymax = mean_dens + se_dens),
                width = 0, color = "black") +
  geom_text(data = raw_summary, aes(x = UV_mean, y = mean_dens, label = area_habitat),
            vjust = -1, size = 3) +
  labs(x = "UV-B (Jan–Aug mean)", y = "Mean density",
       title = "Predicted trend vs. raw per-habitat means") +
  theme_cowplot()
```


```{r}
base_colors <- c(
  "Madrid"     = "#E63946",
  "Lisbon"     = "#2A9D8F",
  "Milan"      = "#457B9D",
  "Malmö"      = "#F4A261",
  "Gothenburg" = "#8E44AD"
)

FeatherAd_db2_UV <- FeatherAd_db2_UV %>%
  mutate(area = recode(area,
                        "Malmo"    = "Malmö",
                        "Goteborg" = "Gothenburg"),
         area = factor(area, levels = names(base_colors)))


summary_db_uv <- FeatherAd_db2_UV %>%
  group_by(area, habitat) %>%
  summarise(
    UV_mean = mean(UV_B_may_au, na.rm = TRUE),
    mean_md = mean(mean_density, na.rm = TRUE),
    se_md   = sd(mean_density, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) %>%
  mutate(fill_color = ifelse(habitat == "Urban", base_colors[as.character(area)], "white"))


plot_UV <- ggplot() +
  geom_ribbon(data = pred, aes(x = x, ymin = conf.low, ymax = conf.high),
              alpha = 0.1, fill = "grey40") +
  geom_line(data = pred, aes(x = x, y = predicted),
            color = "black", linewidth = 1, linetype = "solid") +
  geom_errorbar(data = summary_db_uv, aes(x = UV_mean, ymin = mean_md - se_md, ymax = mean_md + se_md,
                                           color = area),
                width = 0) +
  geom_point(data = summary_db_uv, aes(x = UV_mean, y = mean_md, color = area, fill = fill_color),
             shape = 21, size = 4, stroke = 1.2) +
  scale_color_manual(values = base_colors, name = NULL) +
  scale_fill_identity() +
  labs(
  x = expression(paste("Mean UV-B radiation (J/", m^2, "/day)")),
  y = "Mean barbule density (barbules/mm)"
)+
  scale_y_continuous(limits = c(18, 22.5), breaks = seq(18, 22.5, by = 0.5)) +
  theme_bw(base_size = 20) +
  theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 12),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.minor.y = element_blank()
  ) +
  guides(
    color = guide_legend(
      nrow = 1,
      override.aes = list(
        shape = 21,
        size = 3.5,
        fill  = base_colors[levels(summary_db_uv$area)],
        color = base_colors[levels(summary_db_uv$area)]
      )
    )
  )

habitat_legend <- ggplot(data.frame(habitat = c("Urban", "Rural"),
                                    x = c(3.1, 3.9), y = c(1, 1))) +
  geom_point(aes(x = x, y = y, fill = habitat), 
             shape = 21, size = 3.5, stroke = 1.2, color = "black") +
  geom_text(aes(x = x, y = y, label = habitat), 
            hjust = -0.3, size = 4) +
  scale_fill_manual(values = c("Urban" = "black", "Rural" = "white")) +
  theme_void() +
  theme(legend.position = "none") +
  coord_cartesian(clip = "off") +
  xlim(0, 5) +
  ylim(0.5, 1.5)

p3 <- plot_grid(plot_UV, habitat_legend, ncol = 1, rel_heights = c(1, 0.12))

p3

ggsave("Plots brevia/p3.pdf", plot = p3, width = 7, height = 7, dpi = 600)

library(egg)

p1_fixed <- set_panel_size(p1, width = unit(5, "in"), height = unit(5, "in"))
p2_fixed <- set_panel_size(plot_elevation, width = unit(5, "in"), height = unit(5, "in"))
p3_fixed <- set_panel_size(plot_UV,        width = unit(5, "in"), height = unit(5, "in"))

ggsave("Plots brevia/p3_fixed.pdf", plot = p3_fixed, width = 9, height = 7, dpi = 600)
ggsave("Plots brevia/p2_fixed.pdf", plot = p2_fixed, width = 9, height = 7, dpi = 600)
ggsave("Plots brevia/p1_fixed.pdf", plot = p1_fixed, width = 9, height = 7, dpi = 600)
```


